---
title: "Data Exploration"
author: "Brenwin"
date: "14/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

# read in data
volcano <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/volcano.csv')
eruptions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/eruptions.csv')
events <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/events.csv')
tree_rings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/tree_rings.csv')
sulfur <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/sulfur.csv')

rof <- read_csv("all.csv")
```

# Questions
- Primary:
  Does volcanoes with different VEI have specific characteristics?

- Secondary:
  How worried should you be about a volcano eruption?
--> generally most volcanoes that erupt are not near populations  
--> high VEI eruptions; don't occur frequently

Which tectonic settings have higher or lower VEI and how this relates to the duration of eruption?
--> most erupting volcanoes (regardless of duration) share a common tectonic setting

*which tectonic settings have higher/lower VEI?

What is the ideal setting for a volcano to form?
--> most of the volcanoes that erupt lie on ring of fire  
-> which are not situated near populations

--> order of VEI isn't that significant in explaining duration or frequency of eruption



# ========== Volcano Data set

## About
- Gives information of 958 volcanoes such as
  - major/minor rocks// tectonic settings
  - volcano_name// country// regions// lat// long
  - population within 5,10,30,100km

## Types of Volcanoes
```{r}
#types of volcanoes
volcano %>% 
  count(primary_volcano_type) 
```

## Categorize volcanoes into 4 main types 
```{r}
volcano_types <- volcano %>% 
  # categorize volcanoes into 4 main types
  mutate(primary_volcano_type = case_when(
    str_detect(primary_volcano_type, "Stratovolcano") ~ "Stratovolcano",
    str_detect(primary_volcano_type, "Shield") ~ "Shield",
    str_detect(primary_volcano_type, "Caldera") ~ "Caldera",
    TRUE ~ "Other"
  ))

unique(volcano_types$primary_volcano_type)
```
- 4 Main types of Volcanoes
  - Shield
  - Stratovolcano
  - Caldera
  - Other


## Recategorize volcanoes' elevation

### volcanoes' elevation? What is volcano average elevation
```{r}
# average elevation of volcanoes
volcano %>% 
  ggplot() +
  geom_histogram(aes(x = elevation),
                 binwidth = 500,
                 fill = "red",
                 colour = "black") +
  xlim(range(volcano$elevation)) +
  coord_flip()

# by volcano types
volcano_types %>% 
  ggplot(aes(x = fct_reorder(primary_volcano_type, elevation),
             y = elevation,
             fill = fct_reorder(primary_volcano_type, elevation))) +
  geom_boxplot() +
  coord_flip() +
  theme(legend.position = "none") +
  geom_point(data = volcano_types,
             aes(x = primary_volcano_type,
             y = elevation),
             position = "jitter",
             alpha = 0.4) # make points less clustered in horizontal line (overplotted)
```

- Elevation of volcanoes; mostly around 1,250m - 1,750m.

### Which volcanoes have the highest elevation?
```{r}
volcano_elevation <- volcano_types %>% 
  arrange(-elevation) %>% 
  select(volcano_name, country, primary_volcano_type, elevation) %>% 
  # concatenate volcano_name and country
  mutate(volcano = str_c(volcano_name, country, sep = " -"),
         # keep unused variable i.e. volcano type 
         .keep = "unused") %>% 
  # rearrange column in data frame
  .[,c(3, 1, 2)] 

volcano_elevation
```


## which countries with most volcanoes?
```{r}
volcano %>% 
  count(country) %>% 
  arrange(-n) %>% 
  head(10)
```

## what about Australia?
```{r}
volcano %>% 
  filter(country == "Australia")
```


# ========== Eruption Data set

One row in the eruptions data set represents a case of volcano eruption

## What is the date range of this data set?
```{r}
eruptions %>% 
  summarize(range(start_year, na.rm = TRUE))
```
- This data set dates back all the way to -11,345 years ago to the present.
  - dubbed the Holocene period. i.e. time since the end of the last "ice age" 

## How many times each volcano erupted? *note: `volcano_erupted` data set include only volcanoes that erupted
```{r}
# join only volcanoes that erupted
volcano_erupted <-inner_join(volcano_types, eruptions) %>%
  # select relevant variables
  select(volcano_name, vei, start_year, end_year, last_eruption_year, primary_volcano_type, country, latitude, longitude, elevation, evidence_category, population_within_5_km, population_within_10_km, population_within_30_km, population_within_100_km) %>% 
# concatenate country & volcano name
  mutate(volcano = str_c(volcano_name, country, sep = " -"), .keep = "unused") 

# summarize volcano erupted; keeping all columns
volcano_erupted <- volcano_erupted %>% 
  group_by(volcano) %>%
  # find count (i.e. no. of times volcano erupted)
  mutate(n = n()) %>%  
  distinct(volcano, 
           .keep_all = TRUE) %>% 
  ungroup()

```

- There were total of 11,1178 eruptions during this year. (roughly about 1 per year!!!)
  - Eruptions are caused by 688 volcanoes

### what; mean & median of number of volcano eruptions per volcano?

#### All time
```{r}
volcano_erupted %>% 
  summarise(median = median(n),
            mean = mean(n))
```

## How worried should you be about volcano eruptions?

- Explositivity of volcanoes; measured through VEI (volcano explositivity index)
  - theoretically: scale 0 to infinity
  - largest VEI measured: 8
  - most recent VEI 7 eruption was observed is in 1815: The eruption at Mount Tambora
    
```{r}
unique(eruptions$evidence_method_dating)

eruptions %>% 
  count(evidence_method_dating)
```

```{r}
# filter eruptions; to after 1812: eruption of Mount Tambora
eruptions_1812 <- eruptions %>% 
  filter(start_year >= 1812) 

# see lag; between last eruption to next
eruptions_1812 <- eruptions_1812 %>% 
  arrange(start_year, volcano_name) %>%  
  group_by(volcano_name) %>% 
  # find year between eruptions = year of eruption + (time between last and start of eruption)
  mutate(lag_year = start_year + (lag(start_year) - start_year))

# Plot all volcanoes that erupted more than once
eruptions_1812 %>% 
  # filter volcanoes; with only 1 eruption i.e. NA for lag_year variable
  filter(!is.na(lag_year)) %>% 
  ggplot() +
  geom_tile(aes(x = volcano_name,
                 y = lag_year),
            size = 3,
            colour = "red") +
  coord_flip() +
  # facet by VEI
  facet_wrap(~vei)
```

## probability distribution function of volcano eruptions in each VEI
```{r}
library(ggridges)
eruptions_1812 %>% 
  filter(!is.na(vei)) %>% 
  group_by(volcano_name, vei) %>% 
  mutate(no_of_eruptions = n(),
         vei = as.factor(vei)) %>%
  ggplot() +
  stat_density_ridges(aes(y = vei,
                          x = no_of_eruptions,
                          fill = vei),
                      quantile_lines = TRUE,
                      quantiles = 0.5,
                      alpha = 0.7,
                      jittered_points = TRUE,
                      position = "points_jitter") +
  scale_x_continuous(breaks = seq(from = 0, to = 70, by = 5 ))
```
- The probability of an eruption of VEI 5 and above is actually very rare



# ========== World map
```{r}
# Plot world map
world <- map_data("world")

world %>% 
  ggplot() +
  geom_map(map = world,
           aes(x = long, y = lat,
               map_id = region)) +
  geom_point(data = volcano_types,
             aes(x = longitude, y = latitude,
                 colour = primary_volcano_type,
                 alpha = 0.7,
                 size = elevation)) +
  theme(legend.position = "bottom") 



```

- where is the ring of fire?
- edit elevation sequence  

### Plot a world map of where volcanoes erupt

```{r}
rof_longfixed <- rof %>% 
  mutate(longfixed = case_when(lon < 0 ~ "neg_long",
                               lon > 0 ~ "pst_long"))

rof_neg_long <- rof_longfixed %>% 
  filter(longfixed == "neg_long")

rof_pst_long <- rof_longfixed %>% 
  filter(longfixed == "pst_long")
```

```{r}
# add continent coordinates
continent_coord <- tibble(id = c("Africa", "Asia", "Europe", "North America", "Oceania", "South America"),
                          long = c(15, 80, 20, -100, 150, -60),
                          lat = c(15, 35, 50, 40, -25, -15),
                          stringsAsFactors = F)

# ----- plot world map
world %>% 
  ggplot() +
  geom_map(map = world,
           aes(x = long, y = lat,
               map_id = region)) +
  geom_point(data = volcano_erupted,
             aes(x = longitude, y = latitude,
                 colour = "red",
                 size = n)) +
  theme(legend.position = "bottom") +
  geom_polygon(data = rof,
               aes(x = lon,
                   y = lat,
                   group = plate),
               fill = NA,
               colour = "dark green") +
  # add continent text
  geom_text(data = continent_coord,
            aes(x = long, 
                y = lat,
                label = id),
            colour = "blue",
            size = 5) +
  # remove legends
  guides(colour = FALSE,
         size = FALSE,
         alpha = FALSE)
```
`


### Plot world map of countries fill with highest no. of volcanoes
```{r}
# clean volcano$country to match world data set
volcano_country <- volcano %>% 
  mutate(country = case_when(
    country == "Armenia-Azerbaijan" ~ "Armenia",
    country == "Burma (Myanmar)" ~ "Burma",
    country %in% c("Chile-Argentina","Chile-Bolivia", "Chile-Peru") ~ "Chile",
    country == "China-North Korea" ~ "China",
    country == "Colombia-Ecuador" ~ "Colombia",
    country %in% c("DR Congo", "DR Congo-Rwanda") ~ "Democratic Republic of the Congo",
    country == "El Salvador-Guatemala" ~ "El Salvador",
    country == "Eritrea-Djibouti" ~ "Eritrea",
    country %in% c("Ethiopia-Djibouti", "Ethiopia-Kenya") ~ "Ethiopia",
    country == "Guatemala-El Salvador" ~ "Guatemala",
    country == "Japan - administered by Russia" ~ "Japan",
    country == "Mexico-Guatemala" ~ "Mexico",
    country == "North Korea-South Korea" ~ "North Korea",
    country == "Saint Kitts and Nevis" ~ "Saint Kitts",
    country == "Syria-Jordan-Saudi Arabia" ~ "Syria",
    country == "Saint Vincent and the Grenadines" ~ "Saint Vincent",
    country == "Uganda-Rwanda" ~ "Uganda",
    country == "United Kingdom" ~ "UK",
    country == "United States" ~ "USA",
    TRUE ~ country
  )) 

volcano_country_count <- volcano_country %>% 
  group_by(country) %>% 
  count() 

volcano_country_join <- left_join(world, volcano_country_count,
          by = c("region" = "country")) %>% 
  # replace n = NA to 0
  mutate(n = ifelse(is.na(n), 0, n)) <- 
```


# ========== events data set

## when; started documenting events?
```{r}
range(events$event_date_year, na.rm = TRUE)
```
- since 9650 years ago.

## What sort of events were documented?
```{r}
unique(events$event_type)
```






