---
title: "ETC5521 Assignment 1"
subtitle: "Your title"
team: Possum
author:
  - Brenwin Ang
  - Joyce Lee
date: "`r Sys.Date()`"
output: 
  html_document
---

[This assignment is for ETC5521 Assignment 1 by Team `r rmarkdown::metadata$team` comprising of `r knitr::combine_words(rmarkdown::metadata$author)`.]{style="color:#006DAE;"}

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-libraries, include = FALSE}
library(tidyverse)
```

```{r, eval = FALSE}
# ----- cleaning script 
library(readxl)
library(tidyverse)

eruption_list <- read_csv("2020/2020-05-12/eruption_list.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  select(-contains("modifier"), -contains("uncertainty"))

event_list <- read_csv("2020/2020-05-12/event_list.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  select(-contains("modifier"), -contains("uncertainty"))

volcano_list <- read_csv("2020/2020-05-12/volcano_list.csv", skip = 1) %>% 
  janitor::clean_names()

eruption_list %>% 
  write_csv("2020/2020-05-12/eruptions.csv")

event_list %>% 
  write_csv("2020/2020-05-12/events.csv")

volcano_list %>% 
  write_csv("2020/2020-05-12/volcano.csv")
```

```{r getting-data}
# Get the Data

volcano <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/volcano.csv')
eruptions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/eruptions.csv')
events <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/events.csv')
tree_rings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/tree_rings.csv')
sulfur <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/sulfur.csv')

# --- Or read in with tidytuesdayR package (https://github.com/thebioengineer/tidytuesdayR)
# PLEASE NOTE TO USE 2020 DATA YOU NEED TO USE tidytuesdayR version ? from GitHub

# Either ISO-8601 date or year/week works!

# Install via devtools::install_github("thebioengineer/tidytuesdayR")

# tuesdata <- tidytuesdayR::tt_load('2020-05-12')
# tuesdata <- tidytuesdayR::tt_load(2020, week = 20)

# volcano <- tuesdata$volcano
```


# Introduction and motivation

[FILL] Give the bigger picture of the data, and inspire the reader to learn more about the problem by reading your analysis. 

 
# Data description
The data source is from [The Smithsonian Institution](https://volcano.si.edu/), which has been constantly updating since 2013. The data is cleaned and made available for download on (https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-05-12/readme.md), so that we directly import the 5 data sets for this analysis. These 5 data sets are probably the most comprehensive data set around. They are volcano, eruptions, events, tree_rings and sulfur, allowing us to discover the influences among them. 

- `volcano`: Provides location, geological and population information on 958 volcanoes with 26 variables such as
  - its metadata: volcano name, type, country location, evidence category, longitude and latitude
  - technical details: major & minor rocks, tectonic settings
  - distance from nearest population (km)
  - Missing data in ""last_eruption_year", "major_rock" and "minor_rock", labelled as "unknown" or left as blank

- `eruptions`: Details eruptions occured since Holecene period (11,345 years ago) to present, comprising of 15 variables. Details include  
  - VEI (volcano explosivity index) as an indication for the severity of eruptions in ascending order, taking values from 0 to 7 
  - date of eruption, eruption category/type, evidence method
  - Unknown data appears commonly among variables "area_of_activity", "start_month", "start_day", "end_year", "end_month" and "end_day", denoted with "NA".

- `events`: any event that occur at the volcanoes documented with 10 variables.
  - such as explosion, dome/cone formation, fatalities, evacuation, loud audible noises among others.
  - Despite variables such as "event_type" and "event_remark" , a lot of its variables (e.g. eruption dates) overlap with that of `eruptions`, thus it seems as a condensed version of `eruptions`.
  - "NA" are presented regularly for "event_remarks" and mostly event dates information.

- `tree_ring`: Tree rings were used as a climate proxy. In study of effects on volcanoes on climate change, researchers matched effects of eruptions to tree ring records. The measurements are conducted yearly and no data is missing.

- `sulfur`
  - Provide information on sulphur detected from 500.CE to 706.CE with 3 variables.
  - Using the same unit for measurement, the data for "neem" is collected from Greenland while it is from Anatartica for "wdc".

# Research questions
- Primary:
  Does volcanoes with different VEI have specific characteristics?

- Secondary:
  How worried should you be about a volcano eruption?
  Which tectonic settings have higher or lower VEI and how this relates to the duration of eruption?
  What is the ideal setting for a volcano to form?
# Analysis and findings

[FILL] Should include at least one plot or numerical summary for each of your questions, that helps the reader arrive at an answer. You should also write paragraphs describing the methods, summaries and findings. 

# References
