---
title: "ETC5521 Assignment 1"
subtitle: "Your title"
team: Possum
author:
  - Brenwin Ang
  - Joyce Lee
date: "`r Sys.Date()`"
output: 
  html_document
---

[This assignment is for ETC5521 Assignment 1 by Team `r rmarkdown::metadata$team` comprising of `r knitr::combine_words(rmarkdown::metadata$author)`.]{style="color:#006DAE;"}

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-libraries, include = FALSE}
library(tidyverse)
```

```{r, eval = FALSE}
# ----- cleaning script 
library(readxl)
library(tidyverse)

eruption_list <- read_csv("2020/2020-05-12/eruption_list.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  select(-contains("modifier"), -contains("uncertainty"))

event_list <- read_csv("2020/2020-05-12/event_list.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  select(-contains("modifier"), -contains("uncertainty"))

volcano_list <- read_csv("2020/2020-05-12/volcano_list.csv", skip = 1) %>% 
  janitor::clean_names()

eruption_list %>% 
  write_csv("2020/2020-05-12/eruptions.csv")

event_list %>% 
  write_csv("2020/2020-05-12/events.csv")

volcano_list %>% 
  write_csv("2020/2020-05-12/volcano.csv")
```

```{r getting-data}
# Get the Data

volcano <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/volcano.csv')
eruptions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/eruptions.csv')
events <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/events.csv')
tree_rings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/tree_rings.csv')
sulfur <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/sulfur.csv')

# --- Or read in with tidytuesdayR package (https://github.com/thebioengineer/tidytuesdayR)
# PLEASE NOTE TO USE 2020 DATA YOU NEED TO USE tidytuesdayR version ? from GitHub

# Either ISO-8601 date or year/week works!

# Install via devtools::install_github("thebioengineer/tidytuesdayR")

# tuesdata <- tidytuesdayR::tt_load('2020-05-12')
# tuesdata <- tidytuesdayR::tt_load(2020, week = 20)

# volcano <- tuesdata$volcano
```


# Introduction and motivation

[FILL] Give the bigger picture of the data, and inspire the reader to learn more about the problem by reading your analysis. 


 
# Data description

The data source is from [The Smithsonian Institution](https://volcano.si.edu/). The data is supplied and cleaned, downloadable from [tidytuesday github](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-05-12/readme.md).

This Volcano data set is probably the most comprehensive data set around. They contain 5 linked data sets:  

- `volcano`: Provides information on 958 volcanoes such as
  - its metadata: volcano name, type, country location, evidence category, longitude and latitude
  - technical details: major & minor rocks, tectonic settings
  - distance from nearest population (km)
  
- `eruptions`: Details eruptions occured since Holecene (dubbed time since the end of "ice age") period (11,345 years ago) to present. Details include  
  - VEI (volcano explosivity index), date of eruption, eruption category/type, evidence method

- `events`: any event that occur at the volcanoes documented.
  - such as explosion, dome/cone formation, fatalities, evacuation, loud audible noises among others.

- `tree_ring`: Tree rings were used as a climate proxy. In study of effects on volcanoes on climate change, researchers matched effects of eruptions to tree ring records. 


- `sulfur`
  - Provide information on sulphur detected from 500.CE to 706.CE

# Analysis and findings

[FILL] Should include at least one plot or numerical summary for each of your questions, that helps the reader arrive at an answer. You should also write paragraphs describing the methods, summaries and findings. 

# References
